Mechanising Recursion Schemes with Magic-Free Coq Extraction

### 2024-10-10 paper/paper.tex

$(A, c_{1})$ is a recursive coalgebra
$(B, c_{2})$ is a recursive coalgebra

$(C, a_{2})$ is an algebra
$(B, a_{1})$ is an algebra

$c_{2} \circ a_{1} = id$

\begin{equation}
  \label{eq:hylo:deforestation}
  \hylo{a_{2}}{c_{2}} \circ \hylo{a_{1}}{c_{1}} = \hylo{a_{2}}{c_{1}}
\end{equation}

\subsection{Recursive Anamorphisms}
Anamorphisms applied to recursive coalgebras specialise to hylomorphisms into an
inductive data type in the following way. A recursive coalgebra
can be applied only finitely many times, therefore when this is applied to an
anamorphism the only possible traces it can produce from the seed function are
the finite ones. We denote this special kind of recursion scheme \emph{recursive
anamorphism}. We can show that recursive anamorphisms of type $X \to \nu F$
can also be given the type $X \to \mu F$. Moreover, these anamorphisms are
exactly hylomorphisms on the recursive $F$-coalgebra and the
algebra $\InIso$ for the inductive data type $\nu F$. This fact falls out
from the universality property of the hylomorphism and the fact that recursive
anamorphisms satisfy the same equation.

\subsection{Haskell Implementation of Recursion Schemes}
\begin{minted}{haskell}
newtype Fix f = In {inOp :: f (Fix f )}
\end{minted}

\begin{minted}{haskell}
data ListF a x = Nil | Cons a x deriving Functor
data TreeF a x = Empty | Node x a x deriving Functor
\end{minted}

\begin{minted}{haskell}
cata :: Functor f => (f a -> a) -> Fix f -> a
cata alg = alg . fmap (cata alg) . In
\end{minted}
\begin{minted}{haskell}
foldr :: (a -> b -> b) -> b -> [ a ] -> b
foldr f e [ ]    = e
foldr f e (x:xs) = f x (foldr f e xs)
\end{minted}
which is very useful in list processing. As a fold, The frequently used function map is also a fold:
\begin{minted}{haskell}
map :: (a -> b) -> [a] -> [b]
map f = foldr (\x xs -> (f x):xs) []
\end{minted}

 and can be
thought of as the $\operator{fmap}$ program in functional programming.
\begin{minted}{haskell}
class Functor f where
  fmap :: (a -> b) -> f a -> f b
\end{minted}

>>>>

\subsubsection{Haskell Implementation of Hylomorphisms}

which is very useful in list processing. As a fold, The frequently used function map is also a fold:
\begin{minted}{haskell}
map :: (a -> b) -> [a] -> [b]
map f = foldr (\x xs -> (f x):xs) []
\end{minted}
 and can be

\subsubsection{Recursive Anamorphisms}
Anamorphisms applied to recursive coalgebras specialise to hylomorphisms into an
inductive data type in the following way. A recursive coalgebra
can be applied only finitely many times, therefore when this is applied to an
anamorphism the only possible traces it can produce from the seed function are
the finite ones. We denote this special kind of recursion scheme \emph{recursive
anamorphism}. We can show that recursive anamorphisms of type $X \to \nu F$
can also be given the type $X \to \mu F$. Moreover, these anamorphisms are
exactly hylomorphisms on the recursive $F$-coalgebra and the
algebra $\InIso$ for the inductive data type $\nu F$. This fact falls out
from the universality property of the hylomorphism and the fact that recursive
anamorphisms satisfy the same equation.
